\name: Deterministic and Coherent Implicits

### 2017-02-11 src/Translation.tex
more of Phil's notes addressed

In this section we explain the dynamic semantics of $\ourlang$ in terms
of System F's dynamic semantics, by means of a type-directed translation. 
This translation turns implicit contexts into explicit parameters and
statically resolves all queries, much like Wadler and Blott's dictionary
passing translation for type classes~\cite{adhoc}. 
The advantage of this approach is that we simultaneously provide a meaning to
well-typed $\ourlang$ programs and an effective implementation that resolves
all queries statically.

@@

%-------------------------------------------------------------------------------
\subsection{Type-Directed Translation}
Figure~\ref{fig:type} presents the translation rules that convert $\ourlang$
expressions into ones of System F. 
The gray parts of the figure extend the type system with the necessary
information for the translation.

@@

The function $|\cdot|$ takes 
$\ourlang$ types $\rulet$ to System F types T: 
\begin{eqnarray*}
|\alpha| & = & \alpha \\
|\rulet_1 \arrow \rulet_2| & = & |\rulet_1| \arrow |\rulet_2| \\
|\forall \alpha. \rulet| & = & \forall \alpha. |\rulet| \\
|\rulet_1 \iarrow \rulet_2| & = & |\rulet_1| \arrow |\rulet_2| %\\
\end{eqnarray*}
Its reveals that implicit $\ourlang$ arrows are translated to explicit System F function arrows.

@@

The type-directed translation judgment, which extends the typing judgment, is 
\begin{center}
  $\tenv \turns \relation{e}{\rulet}~\gbox{\leadsto E}$
\end{center}
This judgment states that the translation of $\ourlang$ expression $e$ with
type $\rulet$ is System~F expression $E$, with respect to type environment
$\tenv$.

@@

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
\paragraph{Meta-Theory} The type-directed translation of $\ourlang$ to System F exhibits a number
of desirable properties.

\begin{theorem}[Type-preserving translation]\label{thm:type:preservation} Let $e$ be an $\ourlang$
  expression, $\rulet$ be a type, $\tenv$ a type environment and $E$ be a System F expression. If
  $\tenv \turns \relation{e}{\rulet} \leadsto E$, then $|\tenv| \turns \relation{E}{|\rulet|}$.
\end{theorem}
Here we define the translation of the type environment form $\ourlang$ to System F as:
\begin{equation*}
\begin{array}{rcl@{\hspace{2cm}}rcl}
|\epsilon| & = & \epsilon & |\tenv,\alpha| & = & |\tenv|, \alpha \\
|\tenv,x : \rulet| & = & |\tenv|, x : |\rulet| &
|\tenv, \rulet \leadsto x| & = & |\tenv|, x : |\rulet|
\end{array}
\end{equation*}

@@

%-------------------------------------------------------------------------------
\subsection{Dynamic Semantics}
Finally, we define the dynamic semantics of $\ourlang$ as the composition of
the type-directed translation and System F's dynamic semantics.  Following
Siek's notation~\cite{systemfg}, this dynamic semantics is:
\[ \mathit{eval}(e) = V \quad\quad \textit{where } \epsilon \turns \relation{e}{\rulet} \leadsto E \textit{ and } E \rightarrow^* V  \]
with $\rightarrow^*$ the reflexive, transitive closure of System F's standard single-step call-by-value reduction relation (see \cite[Chapter 23]{tapl}).

Now we can state the conventional type safety theorem for $\ourlang$:
\begin{theorem}[Type Safety]
If $\epsilon \turns \relation{e}{\rulet}$, then $\mathit{eval}(e) = V$ for
some System F value $V$.
\end{theorem}
The proof follows trivially from Theorem~\ref{thm:type:preservation}.

>>>>

In this section we explain the dynamic semantics of $\name$ in terms
of System F's dynamic semantics, by means of a type-directed translation. 
This translation turns implicit contexts into explicit parameters and
statically resolves all queries, much like Wadler and Blott's dictionary
passing translation for type classes~\cite{adhoc}. 
The advantage of this approach is that we simultaneously provide a meaning to
well-typed $\name$ programs and an effective implementation that resolves
all queries statically.

@@

%-------------------------------------------------------------------------------
\subsection{Type-Directed Translation}
Figure~\ref{fig:type} presents the translation rules that convert $\name$
expressions into ones of System F. 
The gray parts of the figure extend the type system with the necessary
information for the translation.

@@

The function $|\cdot|$ takes 
$\name$ types $\rulet$ to System F types T: 
\begin{equation*}
\begin{array}{rcl@{\hspace{1cm}}rcl}
|\alpha| & = & \alpha &
|\forall \alpha. \rulet| & = & \forall \alpha. |\rulet| \\
|\rulet_1 \arrow \rulet_2| & = & |\rulet_1| \arrow |\rulet_2| &
|\rulet_1 \iarrow \rulet_2| & = & |\rulet_1| \arrow |\rulet_2| %\\
\end{array}
\end{equation*}
Its reveals that implicit $\name$ arrows are translated to explicit System F function arrows.

@@

The type-directed translation judgment, which extends the typing judgment, is 
\begin{center}
  $\tenv \turns \relation{e}{\rulet}~\gbox{\leadsto E}$
\end{center}
This judgment states that the translation of $\name$ expression $e$ with
type $\rulet$ is System~F expression $E$, with respect to type environment
$\tenv$.

@@

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - 
\paragraph{Meta-Theory} The type-directed translation of $\name$ to System F exhibits a number
of desirable properties.

\begin{theorem}[Type-preserving translation]\label{thm:type:preservation} Let $e$ be an $\name$
  expression, $\rulet$ be a type, $\tenv$ a type environment and $E$ be a System F expression. If
  $\tenv \turns \relation{e}{\rulet} \leadsto E$, then $|\tenv| \turns \relation{E}{|\rulet|}$.
\end{theorem}
Here we define the translation of the type environment form $\name$ to System F as:
\begin{equation*}
\begin{array}{rcl@{\hspace{2cm}}rcl}
|\epsilon| & = & \epsilon & |\tenv,\alpha| & = & |\tenv|, \alpha \\
|\tenv,x : \rulet| & = & |\tenv|, x : |\rulet| &
|\tenv, \rulet \leadsto x| & = & |\tenv|, x : |\rulet|
\end{array}
\end{equation*}

@@

%-------------------------------------------------------------------------------
\subsection{Dynamic Semantics}
Finally, we define the dynamic semantics of $\name$ as the composition of
the type-directed translation and System F's dynamic semantics.  Following
Siek's notation~\cite{systemfg}, this dynamic semantics is:
\[ \mathit{eval}(e) = V \quad\quad \textit{where } \epsilon \turns \relation{e}{\rulet} \leadsto E \textit{ and } E \rightarrow^* V  \]
with $\rightarrow^*$ the reflexive, transitive closure of System F's standard single-step call-by-value reduction relation (see \cite[Chapter 23]{tapl}).

Now we can state the conventional type safety theorem for $\name$:
\begin{theorem}[Type Safety]
If $\epsilon \turns \relation{e}{\rulet}$, then $\mathit{eval}(e) = V$ for
some System F value $V$.
\end{theorem}
The proof follows trivially from Theorem~\ref{thm:type:preservation}.
